---
title: "BIS581v"
output:
  word_document: default
  html_notebook: default
---

BIS 581v Mid project. You have three data sets: central, western and eastern region.  You may need to edit the code below for the path to the files.

```{r}
library(tidyr)
library(tidyverse)
#load from CSV
centralData <- read.csv("Orders_Central.csv", header=TRUE)
westData <- read.csv("orders_west.csv", header=TRUE)
eastData <- read.delim("Orders_East.txt", header=TRUE, sep='\t')
```
For this assignment, answer the following questions, but please note: the data may or may not be setup such that you can just answer. You may need to perform some "data wrangling" before you can answer. Include the code you use for any wrangling/EDA in your submission. Also include the code you use to answer each question along with text indicating what the answer is. You must provide proof of your answer by showing your R code. You may use online resources and talk to others, but what you submit must be your own work.
```{r}
#Wrangling the data
# Combine Year, Month, and Day columns into a Date format
centralData$Order.Date <- as.Date(sprintf("%s-%s-%s", centralData$Order.Year,centralData$Order.Month,centralData$Order.Day))
#The sprintf function in R formats strings with placeholders, allowing precise control over numbers, text, and alignment, 

centralData$Ship.Date <- as.Date(sprintf("%s-%s-%s", centralData$Ship.Year, centralData$Ship.Month,centralData$Ship.Day))

# Convert 'Order.Date' and 'Ship.Date' columns from character to Date, ignoring the time
eastData$Order.Date <- as.Date(eastData$Order.Date, format = "%m/%d/%Y")
eastData$Ship.Date <- as.Date(eastData$Ship.Date, format = "%m/%d/%Y")

# Convert 'Order.Date' and 'Ship.Date' columns to Date, ignoring the time part
westData$Order.Date <- as.Date(substr(westData$Order.Date, 1, 10))
westData$Ship.Date <- as.Date(substr(westData$Ship.Date, 1, 10))

# Set the 'Region' for the central data
centralData$Region <- "Central"

# Remove non-numeric characters from 'Sales' and convert to numeric for calculations
eastData$Sales <- as.numeric(gsub("[^0-9.]", "", eastData$Sales))
#The gsub function in R is used to replace all occurrences of a specified pattern within a string with a replacement value, using regular expressions for pattern matching.


# Combining all data using bind_rows
combined_orders <- bind_rows(centralData, eastData, westData)

```



which region, on average, ships products faster:
```{r}
# Calculate the average shipping time for each region
average_ship_time <- aggregate((combined_orders$Ship.Date - combined_orders$Order.Date), by = list(combined_orders$Region), mean)

# Rename the columns
colnames(average_ship_time) <- c("Region", "Average_Shipping_Time")

# View the result
average_ship_time

# Find the region with the fastest average shipping time
fastest_region <- average_ship_time[which.min(average_ship_time$Average_Shipping_Time), ]
fastest_region




```

Which products ship slowest by region:
```{r}
# Central region: Find the slowest-shipped product
central_slowestproducts <- centralData %>%
  mutate(Shipping_Time = as.numeric(Ship.Date - Order.Date)) %>%
  group_by(Product) %>%
  summarize(slowest_central = max(Shipping_Time, na.rm = TRUE)) %>%
  filter(slowest_central == max(slowest_central))

# East region: Find the slowest-shipped product
east_slowestproducts <- eastData %>%
  mutate(Shipping_Time = as.numeric(Ship.Date - Order.Date)) %>%
  group_by(Product.Name) %>%
  summarize(slowest_east = max(Shipping_Time, na.rm = TRUE)) %>%
  filter(slowest_east == max(slowest_east))

# West region: Find the slowest-shipped product
west_slowestproducts <- westData %>%
  mutate(Shipping_Time = as.numeric(Ship.Date - Order.Date)) %>%
  group_by(Product.Name) %>%
  summarize(slowest_west = max(Shipping_Time, na.rm = TRUE)) %>%
  filter(slowest_west == max(slowest_west))

# Print the slowest-shipped products for each region
print(central_slowestproducts)
print(east_slowestproducts)
print(west_slowestproducts)


```

Plot time to ship by category, by year.
```{r}
# Load required library
library(ggplot2)

# Ensure Order.Date and Ship.Date are in Date format
combined_orders$Order.Date <- as.Date(combined_orders$Order.Date)
combined_orders$Ship.Date <- as.Date(combined_orders$Ship.Date)

# Calculate shipping time in days
combined_orders$Shipping.Time <- as.numeric(difftime(combined_orders$Ship.Date, combined_orders$Order.Date, units = "days"))

# Remove rows with non-finite values in Shipping.Time (e.g., NA or Inf)
combined_orders <- combined_orders[is.finite(combined_orders$Shipping.Time), ]

# Create a new column for the year from Order.Date
combined_orders$Year <- format(combined_orders$Order.Date, "%Y")

# Plot shipping time by category, grouped by year
ggplot(combined_orders, aes(x = Year, y = Shipping.Time)) +
  geom_boxplot() +  # Create boxplot
  facet_wrap(~ Category) +  # Separate plots for each category
  labs(title = "Time to Ship by Category by Year", x = "Year", y = "Shipping Time (Days)")
```

which categories have highest profit by region, chain-wide?
```{r}
# Calculate the total profit for each product category by region
Sum_profit <- aggregate(Profit ~ Region + Category, data = combined_orders, sum)

# Print the total profit for each category by region
print(Sum_profit)

# Find the category with the highest profit chain-wide (across all regions)
highest_chained_profit <- Sum_profit[which.max(Sum_profit$Profit), ]

# Print the category and region with the highest total profit
print(highest_chained_profit)


```

which segments have the lowest profit by region?
```{r}
# Calculate the average profit for each segment by region
Avg_profit <- aggregate(Profit ~ Region + Segment, data = combined_orders, mean)

# Print the average profit for each segment by region
print(Avg_profit)

# Find the segment with the lowest average profit chain-wide (across all regions)
lowest_profit_by_region <- Avg_profit[which.min(Avg_profit$Profit), ]

# Print the segment with the lowest average profit
print(lowest_profit_by_region)


```

What are yearly sales by region?
```{r}
# Convert Order.Date to Date format if not already done
combined_orders$Order.Date <- as.Date(combined_orders$Order.Date)

# Extract the year from Order.Date
combined_orders$Year <- format(combined_orders$Order.Date, "%Y")

# Calculate yearly sales by region
Yearly_sales <- aggregate(Sales ~ Region + Year, data = combined_orders, sum)

# Print the yearly sales for each region
print(Yearly_sales)

```

